---
title: "Lesson 7 Solutions"
author: "Stefanie Molin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "cairo_pdf", warning = FALSE, message = FALSE)
```

```{r credentials, echo = FALSE}
library(getPass)

username <- "s.molin"
password <- getPass("Vertica password:")
```

Let's do some practice problems to challenge your understanding of the company R package and review the material from the prior lessons.

1. Write a query to collect the last 30 days of stats for the client of your choice and save it as a `.sql` file. Make the query simple (like the one we used in the lesson) and dynamic. Use the company R package to read the query into a string and query Vertica once you have modified the query. Use `dplyr` to pivot the data up as you wish, then save the dataframe to a CSV.

```{r solution1_pt1}
# load libraries
library(companyR)
library(dplyr)

# parameters
startDate <- Sys.Date() - 30
client_1_client_id <- 8050

# read query from a file
query <- read_query_from_file("client_stats_query.sql")

# query for last 30 days of client stats for client_1
client_1 <- pull_data_vertica(sprintf(query, startDate, client_1_client_id), cluster = "pa4",
                           username = username, password = password)

# pivot dataframe
client_1_pivot <- client_1 %>%
  select(day, displays, clicks, revenue, pc_conv = post_click_conversions,
         pc_sales = post_click_sales) %>%
  group_by(day) %>%
  summarize(total_clicks = sum(clicks, na.rm = TRUE),
            total_imps = sum(displays, na.rm = TRUE),
            spend = sum(revenue, na.rm = TRUE),
            conv = sum(pc_conv, na.rm = TRUE)) %>%
  mutate(ctr = total_clicks/total_imps, cpc = spend/total_clicks) %>%
  arrange(day)

# see first few rows
head(client_1_pivot, 3)

# write to a CSV
csv_name <- "client_1.csv"
write.csv(x = client_1_pivot, file = csv_name, row.names = FALSE)
```

\pagebreak

2. Using the dataframe you obtained in (1), (a) make a company-themed ggplot showing spend by day. (b) Save the graph as a PNG. (c) Send both the PNG and the CSV from (1) to yourself using the company R package.

```{r solution2_pt1, fig.height=3.7}
# load library
library(ggplot2)

# define aesthetics
(plot <- client_1_pivot %>%
  ggplot(aes(x = as.Date(day), y = spend)) +

  # add line
  geom_line() +

  # format labels
  xlab("Date") +
  scale_y_continuous("Spend", labels = scales::dollar) +

  # add title
  ggtitle("Spend by Day", subtitle = "client_1 L30D") +

  # companyR themes and colors
  companyR::scale_color_company() +
  companyR::theme_company_default())
```

```{r solution2_pt2}
# save the plot as an image
plot_name <- "client_1.png"
ggsave(plot_name, plot)

# send plot and CSV
send_email(username = username, password = password, recipient = "s.molin@company.com",
           attachment_files = c(csv_name, plot_name), subject = "companyR Training Ex 2")
```

\pagebreak

3. Turn the dataframe and ggplot from (1) and (2) into a PowerPoint using the company R package and `ReporteRs`. Send yourself the deck.

```{r solution3}
# create ppt object
ppt <- companyR::ppt_company_template()

# load ReporteRs library
library(ReporteRs)

# add slides and content
# add title slide
ppt <- addSlide(ppt, "main_title")
ppt <- addTitle(ppt, "client_1 Performance Last 30 Days", width = 8)

# add slide for graph
ppt <- ppt %>%
  addSlide("blank_slide") %>%
  addTitle("Spend by Day") %>%
  addPlot(fun = print, x = plot)

# add slide for metrics by day
ppt <- ppt %>%
  addSlide("blank_slide") %>%
  addTitle("Daily Stats") %>%
  addFlexTable(FlexTable(client_1_pivot))

# add an end slide
ppt <- addSlide(ppt, "end_slide")

# save the ppt
filename <- "client_1 Metrics Last 30 Days.pptx"
writeDoc(ppt, file = filename)

# send email with PPT
send_email(username = username, recipient = "s.molin@company.com",
           attachment_files = filename, password = password,
           subject = "client_1 PPT: Solution 3",
           body = "client_1 PPT from companyR Training.")
```
