---
title: "R Training"
subtitle: "Lesson 7"
author: "Stefanie Molin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, dev = "cairo_pdf", warning = FALSE, message = FALSE)
```

```{r credentials, echo = FALSE}
library(getPass)

username <- "s.molin"
password <- getPass("Vertica password:")
```

## I. companyR
You will be happy to know the company has its own R package (`companyR`) that makes it easier to query our databases, make company-themed graphs, generate PowerPoints, and send emails on the company network. The idea behind the package is to curate a collection of functions that everyone at the company can use without having to write their own code for the same task, like querying Vertica. It is constantly being improved, so this document will just be able to highlight the current state of the package.

### A. Installing `companyR`
`companyR` lives on gitlab, not CRAN, so, we need to use a function from the `devtools` package to allow us to install this (`install.packages()` won't work in this case).

* Follow the instructions here: <url>
* Any errors that arise naming a package mean that you need to install that package (i.e. run `install.packages("package")`) and try to install `companyR` again.

### B. `companyR` Functions
Now that we have the companyR package, let's see what functions we have at our disposal. You can see the functions available in a package using: `ls("package:<package_name>")`. If you want to see the arguments for a particular function, but don't need the full help page, use `args(<function_name>)`.

```{r loadcompanyR}
# load companyR
library(companyR)

# show functions in companyR
ls("package:companyR")
```

\pagebreak

```{r args_example}
# see arguments for companyR
args(send_email)
```

### C. `companyR` Use Case
Now, we are going to build on our R knowledge and learn how to use the `companyR` package for various tasks in conjunction with the packages we covered in prior lessons.

#### 1. Querying Databases
Let's pull last 30 days of client data for client_1. We are going to use `companyR` to read the query in from a file and query Vertica. Then we use `dplyr` to pivot our data into `client_1_pivot` which you should remember from the other lessons; the only difference here is how we got the query and the data from Vertica. *Note that `companyR` also has a function to query the SQL server.*

```{r pullData_pt1}
# load dplyr
library(dplyr)

# get date for 30 days ago
startDate <- Sys.Date() - 30

# client_1 client_id
client_1_client_id <- 4624

# read query from a file
query <- read_query_from_file("client_stats_query.sql")

# look at query using stringr to clean up whitespaces for printing
library(stringr)
cat(str_replace_all(query, "\\s{2,}", " "))

# query for last 30 days of client stats for client_1
# (select cluster in function call)
# here we provide the password but if you don't you will be prompted to enter it
client_1 <- pull_data_vertica(sprintf(query, startDate, client_1_client_id), cluster = "pa4",
                           username = username, password = password)
```

\pagebreak

```{r pullData_pt2}
# pivot dataframe and filter down to 25 days
client_1_pivot <- client_1 %>%
  select(day, displays, clicks, revenue, pc_conv = post_click_conversions,
         pc_sales = post_click_sales) %>%
  filter(as.Date(day) >= Sys.Date() - 25) %>%
  group_by(day) %>%
  summarize(total_clicks = sum(clicks, na.rm = TRUE),
            total_imps = sum(displays, na.rm = TRUE),
            spend = sum(revenue, na.rm = TRUE),
            conv = sum(pc_conv, na.rm = TRUE)) %>%
  mutate(ctr = total_clicks/total_imps, cpc = spend/total_clicks) %>%
  arrange(day)

# see first few rows
head(client_1_pivot)
```

#### 2. Company-themed `ggplot2` Graphs
`companyR` has functions for easily formatting your `ggplot2` graphs to company colors. `scale_color_company()` and `scale_fill_company()` will format the colors used in the `col` and `fill` aesthetics, respectively. `theme_company_default()` will handle the rest. *Note that to use this in an RMarkdown PDF you will need to add `dev = "cairo_pdf"` to the `knitr` options at the top of the .Rmd file.*

```{r company_themed_graph, fig.height=4}
# clicks vs. CPC for client_1 colored by days of the week

# load library
library(ggplot2)

# define aesthetics
(plot <- client_1_pivot %>%
  ggplot(aes(x = cpc, y = total_clicks,
             col = factor(format(as.Date(client_1_pivot$day), "%A"),
                          levels = c("Sunday", "Monday", "Tuesday",
                                     "Wednesday", "Thursday", "Friday",
                                     "Saturday")))) +
  # add points
  geom_point() +

  # format labels
  scale_x_continuous("CPC", labels = scales::dollar) +
  scale_y_continuous("Clicks", labels = scales::comma) +

  # add title
  ggtitle("Clicks vs. CPC", subtitle = "client_1 L25D") +

  # companyR themes and colors
  companyR::scale_color_company() +
  companyR::theme_company_default())
```

#### 3. Generate Company-themed PowerPoint Decks
`companyR` provides a function `ppt_company_template()` which will instantiate a PowerPoint object (using the company template slide) which you can use with the `ReporteRs` package to fill with the content of your choosing.

```{r sample_ppt}
# create ppt object
ppt <- companyR::ppt_company_template()

# load ReporteRs library
library(ReporteRs)

# add slides and content
# add title slide
ppt <- addSlide(ppt, "main_title")
ppt <- addTitle(ppt, "client_1 Performance Last 25 Days", width = 8)

# add slide for graph
ppt <- ppt %>%
  addSlide("text_and_xlarge_image_horizontal") %>%
  addTitle("Clicks vs. CPC") %>%
  addParagraph("Here is clicks vs. CPC for the last 25 days.") %>%
  addPlot(fun = print, x = plot)

# add slide for metrics by day
ppt <- ppt %>%
  addSlide("blank_slide") %>%
  addTitle("Daily Stats") %>%
  addFlexTable(FlexTable(client_1_pivot))

# add an end slide
ppt <- addSlide(ppt, "end_slide")

# save the ppt
filename <- "client_1 Metrics Last 25 Days.pptx"
writeDoc(ppt, file = filename)
```

Note that for each slide you create you will need to know the name of that particular layout. You can cycle through the options using the below code:

```{r notes_on_slide_creations, eval=FALSE}
## not run (this is how you see what layouts are available)
# view available slide layouts (by name)
slide.layouts(ppt)

# see how the available slide layouts look
layouts <- slide.layouts(ppt)
for(i in layouts){
	slide.layouts(temp_ppt, i)
	title(sub = i)
	if(interactive()) readline(prompt = "show next slide layout")
}
##
```

#### 4. Sending Emails
You can use the `send_email()` function to send emails (and attachments) using Company credentials.

```{r send_email}
# send the email with the ppt
companyR::send_email(username = username, recipient = "s.molin@company.com",
                    attachment_files = filename, password = password,
                    subject = "client_1 PPT",
                    body = "client_1 PPT from companyR Training.")
```


\pagebreak

## II. Exercises
Let's do some practice problems to challenge your understanding of `companyR` and review the material from the prior lessons.

1. Write a query to collect the last 30 days of stats for the client of your choice and save it as a `.sql` file. Make the query simple (like the one we used in the lesson) and dynamic. Use `companyR` to read the query into a string and query Vertica once you have modified the query. Use `dplyr` to pivot the data up as you wish, then save the dataframe to a CSV.

```{r solution1_pt1}
# load libraries
library(companyR)
library(dplyr)

# parameters
startDate <- Sys.Date() - 30
client_1_client_id <- 8050

# read query from a file
query <- read_query_from_file("client_stats_query.sql")

# query for last 30 days of client stats for client_1
client_1 <- pull_data_vertica(sprintf(query, startDate, client_1_client_id), cluster = "pa4",
                           username = username, password = password)

# pivot dataframe
client_1_pivot <- client_1 %>%
  select(day, displays, clicks, revenue, pc_conv = post_click_conversions,
         pc_sales = post_click_sales) %>%
  group_by(day) %>%
  summarize(total_clicks = sum(clicks, na.rm = TRUE),
            total_imps = sum(displays, na.rm = TRUE),
            spend = sum(revenue, na.rm = TRUE),
            conv = sum(pc_conv, na.rm = TRUE)) %>%
  mutate(ctr = total_clicks/total_imps, cpc = spend/total_clicks) %>%
  arrange(day)

# see first few rows
head(client_1_pivot, 3)

# write to a CSV
csv_name <- "client_1.csv"
write.csv(x = client_1_pivot, file = csv_name, row.names = FALSE)
```

\pagebreak

2. Using the dataframe you obtained in (1), (a) make a Company-themed ggplot showing spend by day. (b) Save the graph as a PNG. (c) Send both the PNG and the CSV from (1) to yourself using `companyR`.

```{r solution2_pt1, fig.height=3.7}
# load library
library(ggplot2)

# define aesthetics
(plot <- client_1_pivot %>%
  ggplot(aes(x = as.Date(day), y = spend)) +

  # add line
  geom_line() +

  # format labels
  xlab("Date") +
  scale_y_continuous("Spend", labels = scales::dollar) +

  # add title
  ggtitle("Spend by Day", subtitle = "client_1 L30D") +

  # companyR themes and colors
  companyR::scale_color_company() +
  companyR::theme_company_default())
```

```{r solution2_pt2}
# save the plot as an image
plot_name <- "client_1.png"
ggsave(plot_name, plot)

# send plot and CSV
send_email(username = username, password = password, recipient = "s.molin@company.com",
           attachment_files = c(csv_name, plot_name), subject = "companyR Training Ex 2")
```

\pagebreak

3. Turn the dataframe and ggplot from (1) and (2) into a PowerPoint using `companyR` and `ReporteRs`. Send yourself the deck.

```{r solution3}
# create ppt object
ppt <- companyR::ppt_company_template()

# load ReporteRs library
library(ReporteRs)

# add slides and content
# add title slide
ppt <- addSlide(ppt, "main_title")
ppt <- addTitle(ppt, "client_1 Performance Last 30 Days", width = 8)

# add slide for graph
ppt <- ppt %>%
  addSlide("blank_slide") %>%
  addTitle("Spend by Day") %>%
  addPlot(fun = print, x = plot)

# add slide for metrics by day
ppt <- ppt %>%
  addSlide("blank_slide") %>%
  addTitle("Daily Stats") %>%
  addFlexTable(FlexTable(client_1_pivot))

# add an end slide
ppt <- addSlide(ppt, "end_slide")

# save the ppt
filename <- "client_1 Metrics Last 30 Days.pptx"
writeDoc(ppt, file = filename)

# send email with PPT
send_email(username = username, recipient = "s.molin@company.com",
           attachment_files = filename, password = password,
           subject = "client_1 PPT: Solution 3",
           body = "client_1 PPT from companyR Training.")
```
