---
title: "Lesson 1 Solutions"
author: "Stefanie Molin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

1. Calculate the average (`mean()`) and standard deviation (`sd()`) of the `speed` column in the `cars` dataframe.
```{r solution_ex1}
mean(cars$speed)
sd(cars$speed)
```

2. Create a new column in the dataframe `cars` called `time` defined as `dist`/`speed`
```{r solution_ex2}
cars$time <- cars$dist/cars$speed
head(cars)
```

\pagebreak

3. A Fibonacci sequence is a series of numbers in which each number is the sum of the two preceding numbers (1, 1, 2, 3, 5, 8, ...). Write a function `fibonacci()` that takes one argument `n`, the size of the sequence you want to print and outputs a Fibonacci sequence of that length. To make this easier, you can assume that the user always properly implements this function (they always provide an n of 1 or greater).

> Extra credit: Have the function handle cases for all numeric values of n and notify the user of an error i.e. if n <= 0 is given.

> *Hint*: Intialize a vector using `numeric(n)` to store your sequence, and have your function *return* the vector.
```{r solution_ex3, error = TRUE}
fibonacci <- function(n){

  # handle missing n input
  if(missing(n)){
    stop("Please provide the length of the sequence you want.")
  }
  
  # handle invalid inputs
  if(n <= 0){
    stop("Not a valid input for n. Values must be greater than or equal to 1.")
  }
  
  # initalize a vector of size n
  fibonacci <- numeric(n)
  
  # handle special cases of n = 1 and 2
  if(n >= 1){
    fibonacci[1] <- 1
    
    if(n >= 2){
      fibonacci[2] <- 1
      
      if(n >= 3){
        # loop through for a series of length n
        for(i in 3:n){
          fibonacci[i] <- fibonacci[i - 2] + fibonacci[i - 1]
        }
      }
    }
  }
  
  return(fibonacci)
}

# check that we can't break the function
fibonacci()
fibonacci(-1)

# check fibonacci on valid inputs
fibonacci(1)
fibonacci(2)
fibonacci(5)
fibonacci(6)
```