---
title: "Client Transaction Analysis"
author: "Stefanie Molin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r QueryVertica_setup, echo = FALSE, warning = FALSE, message = FALSE}
library(RJDBC)
library(getPass)

username <- "s.molin"
password <- getPass("Vertica password:")


# Set .jar file location and Vertica connection string
driverLocation <- "<vertica_driver_path>.jar"
v <- "jdbc:vertica://<vertica_cluster>:<port>/<db>?ConnectionLoadBalance=true"

#' @description Query Vertica for the specified query. Return a dataframe of the results.
#'
#' @param username Vertica login
#' @param query Vertica query (be careful with comments)
#' @param password Vertica password to access the database
#'
#' @return dataframe of results
#'
#' @note Queries are turned into strings so if you comment with "--" the result will be
#' a string with everything after that commented out. Be sure to remove those beforehand.
#'

QueryVertica <- function (username, query, password){
  msg.out <- capture.output(suppressMessages(require(RJDBC)))
  drv <- JDBC("com.vertica.jdbc.Driver", driverLocation)
  conn <- dbConnect(drv, v, username, password)
  data <- dbGetQuery(conn, query)
  dbDisconnect(conn)
  return(data)
}
```

## The Problem

Very senior execs at the client are running into situations where they are seeing purchased products in the banners. We did a test order and found that the product ID didn't fire via the conversion tag in our test order. Given that, is there a way for you to see how many sales didn't have product IDs fire in the *last 30 days*? We're now concerned that we could be showing purchased products in banners **AND** missing out on sales.

## Analysis

We are going to do this in 3 steps:

1. Query Vertica for data
2. Use `dplyr` to manipulate it
3. Graph data using `ggplot2`

### Query Vertica
Pull the following information by day:

* number of transactions with null internal product IDs
* total transactions

```{r query}
# QueryVertica already loaded along with user/pass
query <- "
SELECT
  day
  , COUNT(DISTINCT (CASE WHEN product_internal_id IS NULL
      THEN transaction_id ELSE NULL END)) AS null_id_sales
  , COUNT(DISTINCT transaction_id) AS total_sales
FROM
	transactions_stats_table
WHERE
	day >= CURRENT_DATE() - 30
	AND merchant_id = 5535
GROUP BY
  day
"

data <- QueryVertica(username, query, password)

```

\pagebreak

### Manipulate Data
Use `dplyr` to prepare the data for the graph and for the conclusion; for the graph we need daily data, but for the conclusion we need an overall. Let's also add the percent of null internal ID sales to the total.

```{r pivot, warning=FALSE, message=FALSE}
library(dplyr)

# conclusion data
data_conclusion <- data %>%
  summarize(percent_null = sum(null_id_sales, na.rm = TRUE)/sum(total_sales,
                                                                na.rm = TRUE))

# graph data
data_pivot <- data %>%
  mutate(percent_null = null_id_sales/total_sales, day = as.Date(day)) %>%
  select(day, percent_null) %>%
  arrange(day)

# view graph data
library(knitr)
knitr::kable(head(data_pivot))
```

\pagebreak

### Graph Data
With `ggplot2` create a line graph of this percentage by day.

```{r graph, warning=FALSE, message=FALSE}
library(ggplot2)

data_pivot %>%
  ggplot(aes(x = day, y = percent_null)) +
  geom_line() +
  ggtitle("Percent of Sales with NULL Product ID") + 
  labs(x = "Day", y = "Percent of Sales with NULL Product ID") +
  scale_y_continuous(labels = scales::percent)
```

## Conclusion
The way we store data on our end seems to involve a convoluted process that takes the external product ID from the client finds its internal ID in our system to store in the database; then any time we have the external ID it is from looking in our system not what the client passed. This seems to be the case because I don't see any sales in the last 30 days with an external product ID of NULL, however, I see **`r paste(round(data_conclusion[[1]]*100, 2), "%", sep = "")` of sales** in the *last 30 days* that have had a NULL as the internal ID.
