---
title: "Lesson 5 Solutions"
author: "Stefanie Molin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo = FALSE, warning = FALSE, message = FALSE}
library(RJDBC)
library(getPass)

username <- "s.molin"
password <- getPass("Vertica password:")


# Set .jar file location and Vertica connection string
driverLocation <- "<vertica_driver_path>.jar"
v <- "jdbc:vertica://<vertica_cluster>:<port>/<db>?ConnectionLoadBalance=true"

#' @description Query Vertica for the specified query. Return a dataframe of the results.
#'
#' @param username Vertica login
#' @param query Vertica query (be careful with comments)
#' @param password Vertica password to access the database
#'
#' @return dataframe of results
#'
#' @note Queries are turned into strings so if you comment with "--" the result will be
#' a string with everything after that commented out. Be sure to remove those beforehand.
#'

QueryVertica <- function (username, query, password){
  msg.out <- capture.output(suppressMessages(require(RJDBC)))
  drv <- JDBC("com.vertica.jdbc.Driver", driverLocation)
  conn <- dbConnect(drv, v, username, password)
  data <- dbGetQuery(conn, query)
  dbDisconnect(conn)
  return(data)
}
```

Let's do some practice problems to challenge your understanding.

1. Query Vertica for spend by day for the last 5 days for 3 clients of your choice. Reshape the long dataframe into a wide one with each client as a column.

```{r presolution1, warning=FALSE, message=FALSE}
library(reshape2)
```

```{r solution1_part1}
# QueryVertica has been loaded along with username/password
query <- "
SELECT
	day
	, client_name
	, SUM(revenue) AS spend
FROM
	(SELECT
		client_name
		, client_id
	FROM
		client_lookup_table
	WHERE
		client_name IN ('c1', 'c2', 'c3')
	GROUP BY
		client_name
		, client_id) cl
JOIN
	client_stats_table stats
ON
	cl.client_id = stats.client_id
WHERE
	day >= CURRENT_DATE() - 5
GROUP BY
	day
	, client_name
"

df <- QueryVertica(username, query, password)

# inspect data
head(df)
```

\pagebreak

```{r solution1_part2}
# reshape the data
(reshaped_df <- dcast(melt(df), day ~ client_name + variable))
```

2. Read in a 2M row excerpt of the client catalog from the provided textfile using the `fread()` function from `data.table` (this is faster than base R and automatically detects options). The file will be read into a `data.table`. (a) Drop the `sqlid` column. (b) Rename the `id` column `external_id`. (c) Make the `name` and `external_id` columns keys. (d) Select the `name` and `external_id` of the most expensive item and least expensive item. Limit the name of the selection to 35 characters.

```{r presolution2, warning=FALSE, message=FALSE}
library(data.table)
library(stringr)
```

```{r solution2}
# read in catalog using data.table's fread()
catalog <- fread("client_catalog.txt")

# drop column
catalog <- catalog[, sqlid := NULL]

# update id column name
setnames(catalog, "id", "external_id")

# make keys
setkey(catalog, name, external_id)

# select the most and least expensive items
solution <- catalog[c(which.max(price), which.min(price)), price,
                    by = .(name, external_id)]
solution[, c("max_or_min", "name") := .(c("Max", "Min"), str_sub(name, 1, 35))]
solution
```

3. Using the client catalog you obtained in `(2)`, (a) find number of products with extra data containing the word "promo". (b) Find the unique promo offers and display a few of them. You will need to use a regular expression to find the value in the `extra` field, then you will need to use `str_match()` to find that pattern, and use a function from the `apply` family to get the results of applying that function on all values of `extra`.

> *Hint*: If you are having trouble with the regex, you can take a few entries of the `extra` column in the data.table and work on adapting a regex here: http://regexr.com/. Be sure to look at how `str_match()` works and pick an appropriate `apply` family member; depending on how you do this, you may need to change the type of the object you give the function from the `apply` family.

```{r presolution3, eval=FALSE}
library(stringr)
```

```{r solution3}
# check if each item has "promo" in extra field
on_promo <- lapply(catalog[, extra], str_detect, pattern = "promo")

# calculate number of promo products
sum(unlist(on_promo))

# find all promo offers (this is different than just looking for the word promo)
promos <- apply(as.matrix(catalog[, extra]), 1, FUN = str_match,
                pattern = "(?:promo@V)([^@]+)")[2,]

# remove the NA's from promos
promos <- na.omit(promos)

# find all unique promos
unique_promos <- unique(promos)
length(unique_promos)

# display a few promos
head(unique_promos)
```
