---
title: "Lesson 4 Solutions"
author: "Stefanie Molin"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r QueryVertica_setup, echo = FALSE, warning = FALSE, message=FALSE}
library(getPass)

username <- "s.molin"
password <- getPass("Vertica password:")

# Set .jar file location and Vertica connection string
driverLocation <- "<path_to_your_vertica_driver>.jar"
v <- "jdbc:vertica://<vertica_cluster_here>:<port>/<DB>?ConnectionLoadBalance=true"

#' @description Query Vertica for the specified query. Return a dataframe of the results.
#'
#' @param username Vertica login
#' @param query Vertica query (be careful with comments)
#' @param password Vertica password to access the database
#'
#' @return dataframe of results
#'
#' @note Queries are turned into strings so if you comment with "--" the result will be
#' a string with everything after that commented out. Be sure to remove those beforehand.
#'

QueryVertica <- function (username, query, password){
  msg.out <- capture.output(suppressMessages(require(RJDBC)))
  drv <- JDBC("com.vertica.jdbc.Driver", driverLocation)
  conn <- dbConnect(drv, v, username, password)
  data <- dbGetQuery(conn, query)
  dbDisconnect(conn)
  return(data)
}
```

Let's do some practice problems to challenge your understanding of `ggplot2` and review the material from the prior lessons.

1. Using `ggplot2` create a scatterplot of CPC and clicks for the client of your choice daily for the last 30 days. Add red crosshairs on the graph to indicate points outside the Tukey fence (outliers), which has bounds $Q1 - 1.5 * IQR$ and $Q3 + 1.5 * IQR$ where interquartile range (IQR) is defined as $IQR = Q3 - Q1$ and $Q$ stands for quartile. Note you will need the `quantile()` and `IQR()` functions. Try to write a simple query and use `dplyr` to manipulate your data.

```{r load_packages, message = FALSE, warning = FALSE}
# load libraries
library(dplyr)
library(ggplot2)
library(RJDBC)
```

```{r solution1_pt1}
# QueryVertica() is already sourced and username/password defined

# query for last 30 days of client stats for client_1
query <- "
SELECT
  *
FROM
  client_stats_table
WHERE
  day >= '%s'
  AND client_id = 4624
"

client_1_stats <- QueryVertica(username, sprintf(query, Sys.Date() - 30), password)

# select columns, group by day to summarize, add ctr and cpc and sort by day
client_1_stats_pivot <- client_1_stats %>%
  select(day, clicks, revenue) %>%
  group_by(day) %>%
  summarize(total_clicks = sum(clicks, na.rm = TRUE),
            spend = sum(revenue, na.rm = TRUE)) %>%
  mutate(cpc = spend/total_clicks) %>%
  select(total_clicks, cpc)
```

\pagebreak

```{r solution1_pt2}
# generate plot
client_1_stats_pivot %>%
  ggplot(aes(x = cpc, y = total_clicks)) +
  geom_point() +
  geom_hline(yintercept = quantile(client_1_stats_pivot$total_clicks, 0.25) -
               1.5 * IQR(client_1_stats_pivot$total_clicks), col = "red") +
  geom_hline(yintercept = quantile(client_1_stats_pivot$total_clicks, 0.75) +
               1.5 * IQR(client_1_stats_pivot$total_clicks), col = "red") +
  geom_vline(xintercept = quantile(client_1_stats_pivot$cpc, 0.25) -
               1.5 * IQR(client_1_stats_pivot$cpc), col = "red") +
  geom_vline(xintercept = quantile(client_1_stats_pivot$cpc, 0.75) +
               1.5 * IQR(client_1_stats_pivot$cpc), col = "red") +
  ggtitle("client_1 Clicks vs. CPC Tukey Fence") +
  labs(x = "CPC", y = "Clicks") +
  scale_x_continuous(labels = scales::dollar) +
  scale_y_continuous(labels = scales::comma)
```

\pagebreak

2. Pull clicks and CPC by category by day for the last 30 days on the client of your choice (choose a category level that makes sense). Create a quick base R histogram to see the distribution of CPC. Then, create a scatterplot of CPC and clicks faceted by category (be sure to try both `facet_wrap()` and `facet_grid()` to see which works best for your data). Limit to categories that have more than 1000 clicks on a given day (or a relevant threshold). Add in regression lines. Note that in this example `facet_wrap()` will most likely look the best, but you should still try `facet_grid()`.

```{r solution2_pt1}
# set up dynamic query parameters
category_level <- 2
merchant_id <- 5535

# here we are using %d because we are inserting numbers, but %s will still work
query <- "
SELECT
	name AS category
	, day
	, clicks
	, cpc
FROM
	(SELECT
		category_id
		, day
		, SUM(clicks) AS clicks
		, SUM(revenue)/SUM(clicks) AS cpc
	FROM
		category_stats_table
	WHERE
		category_level = %d
		AND day >= CURRENT_DATE() - 30
		AND merchant_id = %d
	GROUP BY
		category_id
		, day) stats
JOIN
	(SELECT
		category_id
		, name
	FROM
		category_lookup_table
	WHERE
		merchant_id = %d
		AND category_level = %d
	GROUP BY
		category_id
		, name) cat
ON
	cat.category_id = stats.category_id
GROUP BY
	name
	, day
	, clicks
	, cpc
"

client_1_stats_category <- QueryVertica(username,
                               sprintf(query, category_level, merchant_id,
                                       merchant_id, category_level), password)

# use dplyr to filter to categories with more than 1000 clicks per day
client_1_stats_category <- client_1_stats_category %>% filter(clicks > 1000)

# histogram to see distribution of CPC
hist(client_1_stats_category$cpc)
```

\pagebreak

```{r solution2_pt2}
# make ggplot facet_wrap
client_1_stats_category %>%
  ggplot(aes(x = cpc, y = clicks)) +
  geom_point() +
  facet_wrap(~ category) +
  scale_y_continuous(name = "Clicks", labels = scales::comma) +
  scale_x_continuous(name = "CPC", labels = scales::dollar,
                     breaks = scales::pretty_breaks(n = 3)) +
  ggtitle("client_1 Clicks vs. CPC for Categories with > 1000 Clicks in a Day") +
  stat_smooth(method = "lm", se = FALSE, alpha = 0.6)
```

\pagebreak

```{r solution2_pt3}
# make ggplot facet_grid
client_1_stats_category %>%
  ggplot(aes(x = cpc, y = clicks)) +
  geom_point() +
  facet_grid(category ~ .) +
  scale_y_continuous(name = "Clicks", labels = scales::comma) +
  scale_x_continuous(name = "CPC", labels = scales::dollar) +
  ggtitle("client_1 Clicks vs. CPC for Categories with > 1000 Clicks in a Day") +
  stat_smooth(method = "lm", se = FALSE, alpha = 0.6)
```

\pagebreak

3. Graph the percentage of deduplicated sales credited to the company by hour between click and purchase for the last 30 days. If you can't think of a client using the deduplication parameter, use <insert client here>.

> + Write a Vertica query to pull in the amount of deduplicated sales credited to the company by hour between click and purchase as well as the total sales for the last 30 days. Don't bucket the hours in your query--you will work with them in `dplyr`!
+ Using `dplyr` verbs, bucket the hours above 30 to the hour 31 bucket, remove any nonsense data, add a column for percentage of duplicated sales credited to the company, reduce the data into just the 2 columns needed for the graph, and sort the data by bucket. Note it is possible to use all 5 verbs and `group_by()` here and that you should use the pipe operator (`%>%`).
+ Use `ggplot2` to generate a line graph of the dedup ratio by hour.

> *Extra credit: format the y-axis and add labels for the axes and the title.*

```{r solution3}
query <- "
SELECT
	FLOOR((trans_timestamp - click_timestamp)/60/60) AS hours_between
	, COUNT(DISTINCT(CASE WHEN deduplication_matching = 1
  THEN transaction_id ELSE NULL END)) AS deduped_conv
  , COUNT(DISTINCT transaction_id) AS total_conv
FROM
	transactions_stats_table
WHERE
	day >= CURRENT_DATE() - 31
	AND merchant_id = 1749
	AND attribution_type = 'pc'
GROUP BY
	FLOOR((trans_timestamp - click_timestamp)/60/60)"

client_2_stats <- QueryVertica(username, query, password)

client_2_stats %>%
  filter(hours_between >= 0) %>%
  mutate(hour_bucket = ifelse(hours_between > 30, 31, hours_between)) %>%
  group_by(hour_bucket) %>%
  summarize(deduped_conv = sum(deduped_conv, na.rm = TRUE),
            total_conv = sum(total_conv, na.rm = TRUE)) %>%
  mutate(dedup_ratio = deduped_conv/total_conv) %>%
  select(hour_bucket, dedup_ratio) %>%
  arrange(hour_bucket) %>%
  ggplot(aes(x = hour_bucket, y = dedup_ratio)) +
  geom_line() +
  ggtitle("Deduplication Ratio by Hours Between Click and Purchase") +
  labs(x = "hours between click and purchase", y = "deduplication ratio (credited)") +
  scale_y_continuous(labels = scales::percent)
```
